---
title: "Data Management"
output: github_document
---

```{r setup, include = FALSE}

pacman::p_load(
     rio,         # import/export
     here,        # file locator
     purrr,       # iteration
     tidyverse    # data management and visualization
)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

#Load data

pacman::p_load(epirhandbook)

get_data("hospital_linelists.xlsx")

```

## Load Linelist and Simple Loop

```{r}
# import linelist
linelist <- import("./data/linelist_cleaned.rds") 


# Simple loop
for (num in c(1,2,3,4,5)) {  # the SEQUENCE is defined (numbers 1 to 5) and loop is opened with "{"
  print(num + 2)             # The OPERATIONS (add two to each sequence number and print)
}                            # The loop is closed with "}" 


colnames(linelist)
```


# For Loops

Operations

```{r}
# create container to store results - a character vector
cases_demographics <- vector(mode = "character", length = nrow(linelist))

# the for loop
for (i in 1:nrow(linelist)){
  
  # OPERATIONS
  # extract values from linelist for row i, using brackets for indexing
  row_gender  <- linelist$gender[[i]]
  row_age     <- linelist$age_years[[i]]    # don't forget to index!
     
  # combine gender-age and store in container vector at indexed location
  cases_demographics[[i]] <- str_c(row_gender, row_age, sep = ",") 

}  # end for loop


# display first 10 rows of container
head(cases_demographics, 10)
```


Empty vectors, dataframes, lists

```{r}
# Vector
delays <- vector(
  mode = "double",                            # we expect to store numbers
  length = length(unique(linelist$hospital))) # the number of unique hospitals in the dataset

# List
plots <- vector(mode = "list", length = 16)

# DF
delays <- data.frame(matrix(ncol = 2, nrow = 3))
```


Printing

```{r}
# make vector of the hospital names
hospital_names <- unique(linelist$hospital)
hospital_names # print

for (hosp in hospital_names){ 
     hospital_cases <- linelist %>% filter(hospital == hosp)
     print(nrow(hospital_cases))
}
```


Looping plots- Original Plot


```{r}
# create 'incidence' object
outbreak <- incidence2::incidence(   
     x = linelist,                   # dataframe - complete linelist
     date_index = date_onset,        # date column
     interval = "week",              # aggregate counts weekly
     groups = gender,                # group values by gender
     na_as_group = TRUE)             # missing gender is own group

# plot epi curve
plot(outbreak,                       # name of incidence object
     fill = "gender",                # color bars by gender
     color = "black",                # outline color of bars
     title = "Outbreak of ALL cases" # title
     )

```


Looping Plots


```{r}
# make vector of the hospital names
hospital_names <- unique(linelist$hospital)

# for each name ("hosp") in hospital_names, create and print the epi curve
for (hosp in hospital_names) {
     
     # create incidence object specific to the current hospital
     outbreak_hosp <- incidence2::incidence(
          x = linelist %>% filter(hospital == hosp),   # linelist is filtered to the current hospital
          date_index = date_onset,
          interval = "week", 
          groups = gender,
          na_as_group = TRUE
     )
     
     # Create and save the plot. Title automatically adjusts to the current hospital
     plot_hosp <- plot(
       outbreak_hosp,
       fill = "gender",
       color = "black",
       title = stringr::str_glue("Epidemic of cases admitted to {hosp}")
     )
     
     # print the plot for the current hospital
     print(plot_hosp)
     
} # end the for loop when it has been run for every hospital in hospital_names 
```


Tracking Progress of Loop

```{r}
# loop with code to print progress every 100 iterations
for (i in seq_len(nrow(linelist))){

  # print progress
  if(i %% 100==0){    # The %% operator is the remainder
    print(i)

}}
```


## purrr and lists

```{r}
pacman::p_load(
     rio,            # import/export
     here,           # relative filepaths
     tidyverse,      # data mgmt and viz
     writexl,        # write Excel file with multiple sheets
     readxl          # import Excel with multiple sheets
)


```


Import and combine Excel sheets


```{r}
#Create sheet names
sheet_names <- readxl::excel_sheets("./data/hospital_linelists.xlsx")

sheet_names
```


Imports data and assigns each sheet a name


```{r}
combined <- sheet_names %>% 
  purrr::set_names() %>% 
  map(.f = ~import("./data/hospital_linelists.xlsx", which = .x))
```


Combine previous steps with binds rows


```{r}
sheet_names <- readxl::excel_sheets("./data/hospital_linelists.xlsx")  # extract sheet names
 
combined <- sheet_names %>%                                     # begin with sheet names
  purrr::set_names() %>%                                        # set their names
  map(.f = ~import("./data/hospital_linelists.xlsx", which = .x)) %>%  # iterate, import, save in list
  bind_rows(.id = "origin_sheet") # combine list of data frames, preserving origin in new column 
```


Similar example that uses map_at() and c(-1) to skip the first sheet


```{r}
sheet_names <- readxl::excel_sheets("./data/hospital_linelists.xlsx")

combined <- sheet_names %>% 
     purrr::set_names() %>% 
     # exclude the first sheet
     map_at(.f = ~import( "./data/hospital_linelists.xlsx", which = .x),
            .at = c(-1))
```


Split data set and export to separate sheets


```{r}
linelist_split <- linelist %>% 
     group_split(hospital)

view(linelist_split)
```




